Word-Referent Identification Under Multimodal Uncertainty

### 2018-04-26 Writeup/ms.tex
modif ms

  \abstract{Identifying a spoken word in a referential context requires both the
ability to process and integrate multimodal input and the ability to
reason under uncertainty. How do these tasks interact with one another?
We introduce a task that allows us to examine how adults identify words
under joint uncertainty in the auditory and visual modalities. We
propose an ideal observer model which provides an account of how
auditory and visual cues are combined optimally. Model predictions are
tested in three experiments where word recognition is made under two
kinds of uncertainty: category ambiguity and/or distorting noise. In all
cases, the optimal model explains much of the variance in human mean
judgments. In particular, when the signal is not distorted with noise,
participants weight the auditory and visual cues optimally, that is,
according to the relative reliability of each modality. But when one
modality has noise added to it, human perceivers systematically prefer
the unperturbed modality to a greater extent than the optimal model
does. The study provides a formal framework which helps us understand
precisely how word form and word meaning interact in word recognition
under uncertainty. Moreover it offers a first step towards a model that
accounts for form-meaning synergy in early word learning.}
  \keywords{Language understanding; audio-visual processing; word learning; speech
perception; computational modeling. \\

@@

The task of matching the sound to the corresponding visual object has
been extensively studied in the developmental literature since it is
considered to be an crucial instance of early word learning. For
example, many studies focused on how children succeed in this task
despite high referential ambiguity (Medina, Snedeker, Trueswell, \&
Gleitman, 2011; Pinker, 1989; Smith \& Yu, 2008; Suanda, Mugwanya, \&
Namy, 2014; Vlach \& Johnson, 2013; Vouloumanos, 2008; Yurovsky \&
Frank, 2015). However, even when they know the exact meanings of the
words, listeners (both children and adults) often face the task of
recognizing which word the speaker has uttered, especially under noisy
circumstances. The purpose of the current study is to explore the
special case of word recognition under uncertainty when adult observers
have access to multimodal cues from the speech and the referent.

One rigorous way to approach this question is through conducting an
\emph{ideal observer} analysis. This research strategy provides a
characterization of the task/goal and shows what the optimal performance
should be under this characterization.\footnote{It is, thus, a general
  instance of the rational approach to cognition (Anderson, 1990). It
  can also be seen as an instance of Marr's computational level of
  analysis.} When there is uncertainty in the input, the ideal observer
performs an optimal probabilistic inference. For example, in order to
recognize an ambiguous linguistic input, the model uses all available
probabilistic knowledge in order to maximize the accuracy of this
recognition. The ideal observer model can be seen as a theoretical upper
limit on performance. It is not so much a realistic model of human
performance, as much as a baseline against which human performance can
be compared (Geisler, 2003; Rahnev \& Denison, 2018). When there is a
deviation from the ideal, it can reveal extra constraints on human
cognition, such as limitations on the working memory or attentional
resources. The ideal observer analysis has had a tremendous impact not
only on speech related research (Clayards, Tanenhaus, Aslin, \& Jacobs,
2008; Feldman, Griffiths, \& Morgan, 2009; Kleinschmidt \& Jaeger, 2015;
Norris \& McQueen, 2008), but also on many other disciplines in the
cognitive sciences (for reviews, see Chater \& Manning, 2006; Knill \&
Pouget, 2004; Tenenbaum, Kemp, Griffiths, \& Goodman, 2011)

Some of these ideal-observer-based studies are closely related to the
question we are addressing in the current work. For instance, Clayards
et al. (2008) simulated auditory uncertainty by manipulating the
probability distribution of a cue (Voice Onset Time) that differentiated
similar words (e.g., \enquote{beach} and \enquote{peach}). They found
that humans were sensitive to these probabilistic cues and their
judgments closely reflected the optimal predictions. In another work,
Feldman et al. (2009) studied the perceptual magnet effect, which is a
phenomenon that involves reduced discriminability near prototypical
sounds in the native language (Kuhl, 1991). They showed that this effect
can be explained as the consequence of optimally solving the problem of
perception under uncertainty.

@@

On the face of it, the question of combining information from the sound
and the visual referent might seem similar to that of audio-visual
speech integration. Nevertheless, there are at least two fundamental
differences between these two cases, and both can influence the way the
auditory and visual cues are combined: First, in the case of
audio-visual speech, both modalities offer information about the same
underlying speech category. They may differ only in terms of their
informational reliability. In a referential context, however, the
auditory and visual modalities play different roles in the referential
process---in addition to possible differences in informational
reliability. Indeed, the auditory input represents the \emph{symbol}
whereas the visual input represents the \emph{meaning}. It has been
suggested that because of its referential property, speech is a
privileged signal for humans, starting in infancy (see Vouloumanos \&
Waxman, 2014 for a review).\footnote{There is a debate as to whether
  speech is privileged for children and adults for the same reasons.
  Whereas some researchers suggest that speech is privileged for both
  children and adults because of its ability to refer (e.g., Waxman \&
  Markow, 1995), others suggest that speech might \emph{not} have a
  referential status from the start. Rather, speech might be preferred
  by children only because of a low level auditory ``overshadowing''
  (e.g., Sloutsky \& Napolitano, 2003).} Thus, in a referential context,
it is possible that listeners do not treat the auditory and visual
modalities as equivalent sources of information. Instead, there could be
a sub-optimal bias for the auditory modality beyond what is expected
from informational reliability alone.

@@

We investigate how people combine cues from the auditory and the visual
modality to recognize words in a referential context. In particular, we
study how this combination is performed under various degrees of
uncertainty in both the auditory and the visual modality. Imagine, for
example, that someone is uncertain whether they heard \enquote{pea} or
\enquote{bee}, does this uncertainty make them rely more on the referent
(e.g., the object being pointed at)? Vice versa, if they are not sure if
they saw a bee or a fly, does it make them rely more on the sound? More
importantly, when input in both modalities is uncertain to varying
degrees, do they weight each modality according to its relative
reliability (which is the optimal strategy), or do they over-rely on a
particular modality (which is a sub-optimal strategy)?

@@

We introduce a new task that tests word recognition in a referential
context. We use two visual categories (cat and dog) and two auditory
categories (/b/ and /d/ embedded in the minimal pair /aba/-/ada/). For
each participant, an arbitrary pairing is set between the auditory and
the visual categories, leading to two audio-visual word categories
(e.g., dog-/aba/, cat-/ada/). In each trial, participants are presented
with an audio-visual target (the prototype of the target category),
immediately followed by an audio-visual test stimulus
(Figure~\ref{fig:task}). The test stimulus may differ from the target in
both the auditory and the visual components. After these two
presentations, participants press \enquote{same} or \enquote{different.}

@@

This paradigm is adapted from a previous task (Sloutsky \& Napolitano,
2003), which has been used with both children and adults to probe
audio-visual encoding (see Robinson \& Sloutsky, 2010 for a review). In
the testing phase of the original task, participants are asked whether
or not the two audio-visual presentations are \emph{identical}. In the
current study, we are interested, rather, in the categorization, i.e.,
determining whether or not two similar tokens are members of the same
phonological/semantic category. Therefore, testing in our task is
category-based: Participants are asked to press \enquote{same} if they
think the second item (the test) belongs to the same category as the
first (target) (e.g., dog-/aba/), even if there is a slight difference
in the sound, in the referent, or in both. They are instructed to press
\enquote{different} only if they think that the second stimulus was an
instance of the other category (cat-/ada/). The task also includes
trials where pictures are hidden (audio-only) or where sounds are muted
(visual-only). These unimodal trials provide us with the participants'
evaluation of the probabilistic information present in the auditory and
visual categories. As we shall see, these unimodal distributions are
used as inputs to the optimal cue combination model.

@@

We construct an ideal observer model that combines probabilistic
information from the auditory and visual modalities. In contrast to the
model used in most research on multisensory integration (e.g., Ernst \&
Banks, 2002)---which typically studies continuous stimuli (e.g., size,
location)---the probabilistic information in our case cannot be
characterized with \emph{sensory noise}, only. Indeed, our task involves
responses over categorical variables (phonemes and concepts), and
therefore, the optimal model should take into account, not only the
noise variability around an individual perceptual estimate, but also its
\emph{categorical variability}, i.e., the uncertainty related to whether
this perceptual estimate belongs to a given category (see also
Bankieris, Bejjanki, \& Aslin, 2017; Bejjanki et al., 2011). In what
follows, we describe a model that accounts for both type of variability.
First, we describe the model in the simplified case of categorical
variability only. Second, we augment this simplified model to account
for sensory noise.

@@

\noindent This assumption says that, given a word-object mapping, e.g.,
\(W=\)(\enquote{cat}-CAT), variation in the way \enquote{cat} is
pronounced does not correlate with changes in any visual property of the
object CAT, which is a valid
assumption.\footnote{Note that this assumptions is more adequate in the case of arbitrary associations such as ours, and less so in the case of redundant association such as audio-visual speech. In the latter, variation in the pronunciation is expected to correlate, at least to some extent, with lip movements.}

@@

The parameter \(b\) represents the differential between the categories'
prior probabilities. However, since the identity of word categories is
randomized across participants, \(b\) measures, rather, a response bias
to \enquote{same} if \(b > 0\), and a response bias to
\enquote{different} if \(b < 0\). We expect a general bias towards
answering \enquote{different} because of the categorical nature of our
same-different task: When two items are ambiguous but perceptually
different, participants might have a slight preference for
\enquote{different} over \enquote{same}. As for the means, their values
are fixed, and they correspond to the most typical tokens in our
stimuli. Finally, observations from each modality (\(a\) and \(v\)) are
weighted in Equation 1 according to their reliability:
\[\beta_a \propto \frac{1}{\sigma^2_{A}}\]
\[\beta_v \propto \frac{1}{\sigma^2_{V}}.\]

@@

So far, we only accounted for categorical variability. For instance, if
the speaker generates a target production \(a_t\) from an auditory
category \(p(a_t | A) \sim N(\mu_{A}, \sigma^2_{A})\), the ideal model
assumes that it has direct access to this production token (i.e.,
\(a=a_t\)), and that all uncertainty is about the category membership of
this token. However, we might also want to account for internal noise in
the brain and/or external noise in the environment. For example, the
observer might not have access to the exact produced target, but only to
the target perturbed by noise. If we assume this noise to be normally
distributed, that is, \(p(a | a_t) \sim N(a_t, \sigma^2_{N_A})\), then
integrating over \(a_t\) leads to the following simple expression:
\[ p(a | A) \sim N(\mu_{A}, \sigma^2_{A}+\sigma^2_{N_A})\] Similarly, in
the case of sensory noise in the visual modality, we get
\[ p(a | V) \sim N(\mu_{V}, \sigma^2_{V}+\sigma^2_{N_V})\] Finally,
using exactly the same derivation as above, we end up with the following
multimodal weighting scheme in the optimal combination model (Equation
1) which takes into account both categorical and sensory variability:

@@

Equation 1 provides the optimal model's predictions for how
probabilities that characterize uncertainty in the auditory and the
visual modalities can be combined to make categorical decisions.
Parameters' estimates of the probability distributions in each modality
are derived by fitting unimodal posteriors to the participants'
responses in the unimodal conditions, i.e., the condition where only the
sounds are heard or only the pictures are seen
(Figure~\ref{fig:task}).\footnote{Further technical detail about model fitting in the unimodal conditions will be given in the method section of Experiment 1}
Using these derived parameters, the optimal model makes predictions
about responses in the bimodal condition where participants both hear
the sounds and see the pictures.

@@

Comparison of the optimal and the descriptive models allows us, not only
to quantify how much people deviate from optimality, but also to
understand precisely the nature of this deviation. Let \(\sigma^2_{A}\)
and \(\sigma^2_{V}\) be the values of the variances used in the optimal
model (derived from the unimodal conditions), and \(\sigma^2_{Ab}\) and
\(\sigma^2_{Vb}\) be the values observed through the descriptive model
in the bimodal condition. Deviation from optimality is measured in two
ways. First, we measure the change in the values of the variance
specific to each modality, that is, how \(\sigma^2_{A}\) compares to
\(\sigma^2_{Ab}\), and how \(\sigma^2_{V}\) compares to
\(\sigma^2_{Vb}\). Second, we measure changes in the proportion of the
visual and auditory variances, i.e., we examine how
\(\frac{\sigma^2_{A}}{\sigma^2_{V}}\) compares to
\(\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}}\). The first measure allows us to
test if response precision changes for each modality when we move from
the unimodal to the bimodal conditions. The second allows us to test the
extent to which the weighting scheme follows the prediction of the
optimal model. The reason we used the proportion of the variances as a
measure of cross-modal weighting is because this proportion corresponds
to the slope\footnote{Or more precisely the absolute value of the slope}
of the decision threshold in the audio-visual space
(Figure~\ref{fig:model}). The decision threshold is defined as the set
of values in this audio-visual space along which the posterior is equal
to 0.5. Formally speaking, the decision threshold has the following
form:

@@

We compared these models to human responses in three experiments. In
Experiment 1, we studied the case where bimodal uncertainty was due to
categorical variability, only. In Experiment 2 and 3 we added noise in
the background on top of categorical variability.

@@

In this Experiment, we test the predictions of the model in the case
where uncertainty is due to categorical variability only (i.e.,
ambiguity in terms of category membership). We do not add any external
noise to the background and we assume that internal sensory noise is
negligible compared to categorical variability
(\(\sigma^2_{A} \gg \sigma^2_{N_A}\), and
\(\sigma^2_{V} \gg \sigma^2_{N_V}\)). Thus, we use the following cue
weighting scheme:

@@

The first part of the experiment trained participants using only the
prototype pictures and the prototype sounds (12 trials, 4 each from the
bimodal, audio-only, and visual-only conditions). After completing
training, we instructed participants on the structure of the task and
encouraged them to base their answers on both the sounds and the
pictures (in the bimodal condition). There were a total of 25 possible
combinations in the bimodal condition, and 5 in each of the unimodal
conditions. Each participant saw each possible trial twice, for a total
of 70 trials/participant. Trials were blocked by condition and blocks
were presented in random order. The experiment lasted around 15 minutes.

@@

\paragraph{Unimodal condition}\label{unimodal-condition}

Remember that data in this conditions allow us to derive the variances
of both the auditory and the visual categories, and that these variances
are used to make predictions about bimodal data (in the visual and
auditory baselines as well as in the optimal model). These individual
variances were derived as follows (we explain the derivation for the
auditory-only case, but the same applies for the visual-only case). We
use the same Bayesian reasoning as we did in the derivation of the
bimodal model: When presented with an audio instance \(a\), the
probability of choosing the sound category 2 (that is, to answer
\enquote{different}) is the posterior probability of this category
\(p(A_2|a)\). If we assume that both sound categories have equal
variances, the posterior probability reduces to:

@@

Although the paradigm is within-subjects, we did not have enough
statistical power to fit a different model for each individual
participant. Instead, models were constructed with data collapsed across
all participants. That being said, the distribution of responses from
individual participants will also be analyzed. The fit was done with a
nonlinear least squares regression using the NLS package in R (Bates \&
Watts, 1988). We computed the values of the parameters, as well as their
95\% confidence intervals, through non-parametric bootstrap (using 10000
iterations).

@@

\subsubsection{Unimodal conditions}\label{unimodal-conditions}

Average categorization judgments and best fits are shown in
Figure~\ref{fig:unimodal}. The categorization function of the auditory
condition was slightly steeper than that of the visual condition,
meaning that participants perceived the sound tokens slightly more
categorically and whih higher certainty than they did with the visual
tokens. For the auditory modality, we obtained the following
values:\footnote{all CIs in the paper are 95\% confidence intervals.}
\(b_A=\) -0.20 {[}0.02, -0.38{]} and \(\sigma^2_A=\) 2.04 {[}1.66,
2.53{]}. For the visual modality, we obtained \(b_V=\) -0.12 {[}0.06,
-0.28{]} and \(\sigma^2_V=\) 3.33 {[}2.83, 3.92{]}.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{ms_files/figure-latex/unimodal-1} \caption{Human responses in the unimodal conditions. Points represent the proportion of `different' to `same' responses in the auditory-only condition (left), and visual-only condition (right). Error bars are 95\% confidence intervals. Solid lines represent best fits.}\label{fig:unimodal}
\end{figure}

@@

Figure~\ref{fig:bimodal} compares the predictions of the normative
models against human responses. The visual, auditory and optimal model
explained, respectively, 30\%, 67\%, and 89\% of total variance in mean
responses.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{ms_files/figure-latex/bimodal-1} \caption{Human responses vs. models' predictions in the bimodal condition. Shape represents auditory distance from the target, and color represents visual distance from the target.}\label{fig:bimodal}
\end{figure}

@@

As a second analysis, we asked whether the observed variance in the
individual distribution was due to mere sampling errors or whether it
corresponded to a real between-subject variability. We simulated
individual responses from the posterior distribution whose parameters
were fit to the population as a whole (i.e., the descriptive posterior).
The resulting distribution is shown in Figure~\ref{fig:individual}. For
ease of comparison, the simulated distribution was superimposed to the
real distribution. We found that the real distribution had a standard
deviation of \(sd=\) 2.24 which was larger than that of the simulated
distribution (\(sd=\) 1.16), indicating that there was real
between-subject variation beyond sampling errors. This finding means
that the participants varied in terms of how they weighted modalities:
Compared to the predictions of the population-level model, some
participants relied more on the auditory modality, whereas others relied
more on the visual modality.

>>>>

  \abstract{Identifying a spoken word in a referential context requires both the
ability to integrate multimodal input and the ability to reason under
uncertainty. How do these tasks interact with one another? We introduce
a task that allows us to examine how adults identify words under joint
uncertainty in the auditory and visual modalities. We propose an ideal
observer model which provides an account of how auditory and visual cues
are combined optimally. Model predictions are tested in three
experiments where word recognition is made under two kinds of
uncertainty: category ambiguity and/or distorting noise. In all cases,
the optimal model explains much of the variance in human mean judgments.
In particular, when the signal is not distorted with noise, participants
weight the auditory and visual cues optimally, that is, according to the
relative reliability of each modality. But when one modality has noise
added to it, human perceivers systematically prefer the unperturbed
modality to a greater extent than the optimal model does. The study
provides a formal framework which helps us understand precisely how word
form and word meaning interact in word recognition under uncertainty.
Moreover it offers a first step towards a model that accounts for
form-meaning synergy in early word learning.}
  \keywords{Language understanding; audio-visual processing; word learning; speech
perception; computational modeling. \\

@@

The task of matching the sound to the corresponding visual object has
been extensively studied in the developmental literature since it is
considered to be an crucial instance of early word learning. For
example, many studies focused on how children might succeed in this task
despite high referential ambiguity (Medina, Snedeker, Trueswell, \&
Gleitman, 2011; Pinker, 1989; Smith \& Yu, 2008; Suanda, Mugwanya, \&
Namy, 2014; Vlach \& Johnson, 2013; Vouloumanos, 2008; Yurovsky \&
Frank, 2015). However, even when they know the exact meanings of the
words, listeners (both children and adults) often face the task of
recognizing which word the speaker utters, especially under noisy
circumstances (Mattys, Davis, Bradlow, \& Scott, 2012; Peelle, 2018).
The purpose of the current study is to explore novel word recognition by
adults under multimodal uncertainty. We focus on the special case where
people have access to multimodal cues from the auditory speech and the
visual referent.

One rigorous way to approach this question is through conducting an
\emph{ideal observer} analysis. This research strategy provides a
characterization of the task/goal and shows what the optimal performance
should be under this characterization.\footnote{It is, thus, a general
  instance of the rational approach to cognition (Anderson, 1990). It
  can also be seen as an instance of Marr's computational level of
  analysis (Marr, 1982).} When there is uncertainty in the input, the
ideal observer performs an optimal probabilistic inference. For example,
in order to recognize an ambiguous linguistic input, the model uses all
available probabilistic knowledge in order to maximize the accuracy of
this recognition. The ideal observer model can be seen as a theoretical
upper limit on performance. It is not so much a realistic model of human
performance, as much as a baseline against which human performance can
be compared (Geisler, 2003; Rahnev \& Denison, 2018). When there is a
deviation from the ideal, it can reveal extra constraints on human
cognition, such as limitations on the working memory or attentional
resources. The ideal observer analysis has had a tremendous impact not
only on speech related research (Clayards, Tanenhaus, Aslin, \& Jacobs,
2008; Feldman, Griffiths, \& Morgan, 2009; Kleinschmidt \& Jaeger, 2015;
Norris \& McQueen, 2008), but also on many other disciplines in the
cognitive sciences (for reviews, see Chater \& Manning, 2006; Knill \&
Pouget, 2004; Tenenbaum, Kemp, Griffiths, \& Goodman, 2011)

Some of these ideal-observer-based studies are closely related to the
question we are addressing in the current work. For instance, Clayards
et al. (2008) simulated auditory uncertainty by manipulating the
probability distribution of a cue (Voice Onset Time) that differentiated
similar words (e.g., \enquote{beach} and \enquote{peach}). They found
that humans were sensitive to these probabilistic cues and their
judgments closely reflected the optimal predictions. In another work,
Feldman et al. (2009) studied the perceptual magnet effect, a phenomenon
that involves reduced discriminability near prototypical sounds in the
native language (Kuhl, 1991). They showed that this effect can be
explained as the consequence of optimally solving the problem of
perception under uncertainty.

@@

On the face of it, the question of combining information from the sound
and the visual referent might seem similar to that of audio-visual
speech integration. Nevertheless, there are at least two fundamental
differences between these two cases, and both can influence the way the
auditory and visual cues are combined: First, in the case of
audio-visual speech, both modalities offer information about the same
underlying speech category. They may differ only in terms of their
informational reliability. In a referential context, however, the
auditory and visual modalities play different roles in the referential
process---in addition to possible differences in informational
reliability. Indeed, the auditory input represents the \emph{symbol}
whereas the visual input represents the \emph{meaning}. It has been
shown that speech has a privileged status compared to other sensory
stimuli (Edmiston \& Lupyan, 2015; Lupyan \& Thompson-Schill, 2012;
Vouloumanos \& Waxman, 2014; Waxman \& Gelman, 2009; Waxman \& Markow,
1995), and that this privilege is indeed related to the speech's ability
to refer (Waxman \& Gelman, 2009).\footnote{There is, however, a debate
  as to whether speech is privileged for children and adults for similar
  reasons. Whereas some researchers suggest that speech is privileged
  for both children and adults because of its ability to refer (e.g.,
  Waxman \& Gelman, 2009), others suggest that speech might \emph{not}
  have a referential status from the start. Rather, speech might be
  preferred by children only because of a low level auditory
  ``overshadowing'' (e.g., Sloutsky \& Napolitano, 2003).} Thus, in a
referential context, it is possible that listeners do not treat the
auditory and visual modalities as equivalent sources of information.
Instead, there could be a sub-optimal bias for the auditory modality
beyond what is expected from informational reliability alone.

@@

We investigate how people combine cues from the auditory and the visual
modality to recognize words in a referential context. In particular, we
study how this combination is performed under various degrees of
uncertainty in both the auditory and the visual modality. Imagine, for
example, that someone is uncertain whether they heard \enquote{pea} or
\enquote{bee}, does this uncertainty make them rely more on the referent
(e.g., the object being pointed at)? Vice versa, if they are not sure if
they saw a bee or a fly, does it make them rely more on the sound? More
importantly, when input in both modalities is uncertain to varying
degrees, do they weight each modality according to its relative
reliability, which is the optimal strategy, or do they over-rely on a
particular modality, which is a sub-optimal strategy?

@@

We introduce a task adapted from Sloutsky and Napolitano (2003). The
original task has been used with both children and adults to probe
audio-visual encoding (see Robinson \& Sloutsky, 2010 for a review).
Here we use a slightly different version to test word recognition in a
referential context. We use two visual categories (cat and dog) and two
auditory categories (/b/ and /d/ embedded in the minimal pair
/aba/-/ada/). For each participant, an arbitrary pairing is set between
the auditory and the visual categories, leading to two audio-visual word
categories (e.g., dog-/aba/, cat-/ada/). In each trial, participants are
presented with an audio-visual target (the prototype of the target
category), immediately followed by an audio-visual test stimulus
(Figure~\ref{fig:task}). The test stimulus may differ from the target in
both the auditory and the visual components. After these two
presentations, participants press \enquote{same} or \enquote{different.}

@@

In the testing phase of the original task (Sloutsky \& Napolitano,
2003), participants are asked whether or not the two audio-visual
presentations are \emph{identical}. In the current study, we are
interested, rather, in the categorization, i.e., determining whether or
not two similar tokens are members of the same phonological/semantic
category. Therefore, testing in our task is category-based: Participants
are asked to press \enquote{same} if they think the second item (the
test) belongs to the same category as the first (target) (e.g.,
dog-/aba/), even if there is a slight difference in the sound, in the
referent, or in both. They are instructed to press \enquote{different}
only if they think that the second stimulus was an instance of the other
category (cat-/ada/). The task also includes trials where pictures are
hidden (audio-only) or where sounds are muted (visual-only). These
unimodal trials provide us with the participants' evaluation of the
probabilistic information present in the auditory and visual categories.
As we shall see, these unimodal distributions are used as inputs to the
optimal cue combination model.

@@

We construct an ideal observer model that combines probabilistic
information from the auditory and visual modalities. In contrast to the
model used in most research on multisensory integration (e.g., Ernst \&
Banks, 2002)---which typically studies continuous stimuli (e.g., size,
location)---the probabilistic information in our case cannot be
characterized with \emph{sensory noise}, only. Indeed, our task involves
responses over categorical variables (phonemes and concepts), and
therefore, the optimal model should take into account, not only the
noise variability around an individual perceptual estimate, but also its
\emph{categorical variability}, i.e., the uncertainty related to whether
this perceptual estimate belongs to a given category (see also
Bankieris, Bejjanki, \& Aslin, 2017; Bejjanki et al., 2011). In what
follows, we describe a model that accounts for both types of
variability. First, we describe the model in the simplified case of
categorical variability only. Second, we augment this simplified model
to account for sensory noise.

@@

\noindent This assumption says that, given a word-object mapping, e.g.,
\(W=\)(\enquote{cat}-CAT), variation in the way \enquote{cat} is
pronounced does not correlate with changes in any visual property of the
object CAT, which is a valid assumption in the context of our
task.\footnote{Note that this assumptions is more adequate in the case of arbitrary associations such as ours, and less so in the case of redundant association such as audio-visual speech. In the latter, variation in the pronunciation is expected to correlate, at least to some extent, with lip movements.}

@@

The parameter \(b\) represents the differential between the categories'
prior probabilities. However, since the identity of word categories is
randomized across participants, \(b\) measures, rather, a response bias
to \enquote{same} if \(b > 0\), and a response bias to
\enquote{different} if \(b < 0\). We expect a general bias towards
answering \enquote{different} because of the categorical nature of our
same-different task: When two items are ambiguous but perceptually
different, participants might have a slight preference for
\enquote{different} over \enquote{same}. As for the means, their values
are fixed, and they correspond to the most typical tokens in our
stimuli. Finally, observations from each modality (\(a\) and \(v\)) are
weighted in Equation 1 according to their reliability (that is,
according to the \emph{inverse} of their variance):
\[\beta_a \propto \frac{1}{\sigma^2_{A}}\]
\[\beta_v \propto \frac{1}{\sigma^2_{V}}.\]

@@

So far, we only accounted for categorical variability. For instance, if
the speaker generates a target production \(a_t\) from an auditory
category \(p(a_t | A) \sim N(\mu_{A}, \sigma^2_{A})\), the ideal model
assumes that it has direct access to this production token (i.e.,
\(a=a_t\)), and that all uncertainty is about the category membership of
this token. However, we might also want to account for internal noise in
the brain and/or external noise in the environment. For example, the
observer might not have access to the exact produced target, but only to
the target perturbed by noise. If we assume this noise to be normally
distributed, that is, \(p(a | a_t) \sim N(a_t, \sigma^2_{N_A})\), then
integrating over \(a_t\) leads to this new expression of the probability
distribution: \[ p(a | A) \sim N(\mu_{A}, \sigma^2_{A}+\sigma^2_{N_A})\]
Similarly, in the case of sensory noise in the visual modality, we get:
\[ p(a | V) \sim N(\mu_{V}, \sigma^2_{V}+\sigma^2_{N_V})\] Finally,
using exactly the same derivation as above, we end up with the following
multimodal weighting scheme in the optimal combination model (Equation
1) which takes into account both categorical and sensory variability:

@@

Equation 1 provides the optimal model's predictions for how
probabilities that characterize uncertainty in the auditory and the
visual modalities can be combined to make categorical decisions.
Parameter estimates of the probability distributions in each modality
are derived by fitting unimodal posteriors to the participants'
responses in the unimodal conditions, i.e., the condition where only the
sounds are heard or only the pictures are seen
(Figure~\ref{fig:task}).\footnote{Further technical detail about model fitting in the unimodal conditions will be given in the method section of Experiment 1}
Using these derived parameters, the optimal model makes predictions
about responses in the bimodal (i.e., audio-visual) condition where
participants both hear the sounds and see the pictures.

@@

Comparison of the optimal and the descriptive models allows us, not only
to quantify how much people deviate from optimality, but also to
understand precisely the nature of this deviation. Let \(\sigma^2_{A}\)
and \(\sigma^2_{V}\) be the values of the variances used in the optimal
model (derived from the unimodal conditions), and \(\sigma^2_{Ab}\) and
\(\sigma^2_{Vb}\) be the values observed through the descriptive model
in the bimodal condition. Deviation from optimality is measured in two
ways. First, we measure the change in the values of the variance
specific to each modality, that is, how \(\sigma^2_{A}\) compares to
\(\sigma^2_{Ab}\), and how \(\sigma^2_{V}\) compares to
\(\sigma^2_{Vb}\). Second, we measure changes in the proportion of the
visual and auditory variances, i.e., we examine how
\(\frac{\sigma^2_{A}}{\sigma^2_{V}}\) compares to
\(\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}}\). The first measure allows us to
test if response precision changes for each modality when we move from
the unimodal to the bimodal conditions. The second allows us to test the
extent to which the weighting scheme follows the prediction of the
optimal model. The reason we used the proportion of the variances as a
measure of cross-modal weighting is because this proportion corresponds
to the
slope\footnote{Or more precisely the absolute value of the slope.} of
the decision threshold in the audio-visual space
(Figure~\ref{fig:model}). The decision threshold is defined as the set
of values in this audio-visual space along which the posterior is equal
to 0.5. Formally speaking, the decision threshold has the following
form:

@@

We compared these models to human responses in three experiments. In
Experiment 1, we studied the case where bimodal uncertainty was due to
categorical variability, only. In Experiment 2 and 3 we added auditory
and visual noise, respectively, on top of categorical variability.

@@

In this Experiment, we test the predictions of the model in the case
where uncertainty is due to categorical variability only (i.e.,
ambiguity in terms of category membership). We do not add any external
noise to the background and we assume that internal sensory noise is
negligible compared to categorical variability
(\(\sigma^2_{A} \gg \sigma^2_{N_A}\) and
\(\sigma^2_{V} \gg \sigma^2_{N_V}\)). Thus, we use the following cue
weighting scheme:

@@

The first part of the experiment trained participants using only the
prototype pictures and the prototype sounds (12 trials, 4 each from the
bimodal, audio-only, and visual-only conditions). After completing
training, we instructed participants on the structure of the task and
encouraged them to base their answers on both the sounds and the
pictures (in the bimodal condition). There were a total of 25 possible
combinations in the bimodal condition, and 5 in each of the unimodal
conditions. Each participant saw each possible trial twice, for a total
of 70 trials/participant. Trials were blocked by condition and blocks
were presented in random order. The experiment lasted around 15
minutes.\footnote{The experiment can be accessed and played from the
  github repository: \url{https://github.com/afourtassi/WordRec/}}

@@

\paragraph{Unimodal conditions}\label{unimodal-conditions}

Remember that data in these conditions allows us to derive the variances
of both the auditory and the visual categories, and that these variances
are used to make predictions about bimodal data (in the visual and
auditory baselines as well as in the optimal model). These individual
variances were derived as follows (we explain the derivation for the
auditory-only case, but the same applies for the visual-only case). We
use the same Bayesian reasoning as we did in the derivation of the
bimodal model: When presented with an audio instance \(a\), the
probability of choosing the sound category 2 (that is, to answer
\enquote{different}) is the posterior probability of this category
\(p(A_2|a)\). If we assume that both sound categories have equal
variances, the posterior probability reduces to:

@@

Although the paradigm is within-subjects, we did not have enough
statistical power to fit a different model for each individual
participant.\footnote{We had a relatively high number of trials spanning
  all possible audio-visual matchings. Getting enough data points per
  trial per participant would have required running the online
  experimental for a much longer time (more than an hour). This could
  have increased significantly the dropout rate and possibly affected
  the quality of the data collected online.} Instead, models were
constructed with data collapsed across all participants. That being
said, we will also analyze the distribution of individual responses. The
fit was done with a nonlinear least squares regression using the NLS
package in R (Bates \& Watts, 1988). We computed the values of the
parameters, as well as their 95\% confidence intervals, through
non-parametric bootstrap (using 10000 iterations).

@@

Average categorization judgments and best fits are shown in
Figure~\ref{fig:unimodal}. The categorization function of the auditory
condition was slightly steeper than that of the visual condition,
meaning that participants perceived the sound tokens slightly more
categorically and whih higher certainty than they did with the visual
tokens. For the auditory modality, we obtained the following values:
\(b_A=\) -0.20 {[}0.02, -0.38{]} and \(\sigma^2_A=\) 2.04 {[}1.66,
2.53{]}. For the visual modality, we obtained \(b_V=\) -0.12 {[}0.06,
-0.28{]} and \(\sigma^2_V=\) 3.33 {[}2.83, 3.92{]}.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{ms_files/figure-latex/unimodal-1} \caption{Human responses in the unimodal conditions across the three experiments. Points represent the proportion of `different' to `same' responses in the auditory-only condition (left), and visual-only condition (right). Error bars are 95\% confidence intervals. Solid lines represent best unimodal posterior fits.}\label{fig:unimodal}
\end{figure}

@@

Figure~\ref{fig:bimodal} compares the predictions of the normative
models against human responses. Remember that the normative models use
variance estimates from the unimodal conditions (where people see input
from only one modality) to predict data in the bimodal condition (where
people see input from both modalities). Each point represents data form
a particular audio-visual matching (e.g., the visual token \(v\) whose
distance from the target \(v_t\) is \(|v-v_t|=3\), matched with the
auditory token \(a\) whose distance from the target \(a_t\) is
\(|a-a_t|=2\)). The visual, auditory and optimal model explained,
respectively, 30\%, 67\%, and 89\% of total variance in mean responses.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{ms_files/figure-latex/bimodal-1} \caption{Human responses vs. Models' predictions in the bimodal condition across the three experiments. Each point represents data form a particular audio-visual matching. Shape represents auditory distance from the target, and color represents visual distance from the target. Thus, each point is characterized by both  shape and color.}\label{fig:bimodal}
\end{figure}

@@

As a second analysis, we asked whether the observed variance in the
individual distribution was due to mere sampling errors or whether it
corresponded to a real between-subject variability. We simulated
individual responses from the posterior distribution whose parameters
were fit to the population as a whole (i.e., the descriptive posterior).
The resulting distribution is shown in Figure~\ref{fig:individual}. For
ease of comparison, the simulated distribution was superimposed to the
real distribution. We found that the real distribution had a standard
deviation of \(sd=\) 2.24 which was larger than that of the simulated
distribution (\(sd=\) 1.10), indicating that there was real
between-subject variation beyond sampling errors. This finding means
that the participants varied in terms of how they weighted modalities:
Compared to the predictions of the population-level model, some
participants relied more on the auditory modality, whereas others relied
more on the visual modality.

In Experiment 1, we tested word recognition when there was multimodal
uncertainty in terms of category membership only. In real life, however,
tokens can undergo distortions due to noisy factors in the environment
(e.g., car noise in the background, blurry vision in a foggy weather).
In Experiment 2 and 3, we explore this additional level of uncertainty.
